### 2025-07-28 2025/Problems of Informatics/article.tex
translate abstract

The article discusses  the problem of finding the best combination of hyperparameters of machine learning and artificial intelligence methods. In many cases, the efficiency (in a given metric) of the resulting solution for different values of hyperparameters can be quite different.
В таких задачах актуальной является проблема некорректной работы исследуемых artificial intelligence (AI) and machine learning (ML) methods в некоторых (заранее неизвестных) подобластях области изменения гиперпараметров. При этом вычислительная сложность настройки делает невозможным использование ручного или полного перебора. Данные особенности определили необходимость разработки различных методов интеллектуального автоматического подбора гиперпараметров.
С математической точки зрения такая задача может быть представлена как problem of finding a global minimum of a function, given in the form of a ``black box'' and computable only in some part of the search domain. In this case, each computation of the objective function value at some point of the feasible domain may require significant computing resources. The objective function is assumed to satisfy the Lipschitz condition. The existence of subdomains where the objective function is undefined can be interpreted as the existence of some hidden, a priori unknown constraints of the problem. 
The authors propose an approach to solving this type of problem, , which is an extension of the information-statistical global search algorithm (GSA) and takes into account the presence of undefined values of the objective function at some points. В рамках алгоритма происходит разбиение области поиска точками испытаний и оцениваются характеристики подобластей на основе значений целевой функции, вычисленных на их границах. В случае отсутствия информации о значении функции в точке в алгоритме используется предложенная авторами оценка, учитывающая размер исследуемой подобласти. To minimize the number of redundant trials in subdomains where the function is not defined, we used a special method parameter $\alpha$ that regulates the number of trial points in areas of non-computability.

@@

The article provides a detailed description and a diagram of the operation of the modified global search algorithm.
The implementation of the global search algorithm for the case of a not-everywhere computable objective function (GSA-N) was based on the iOpt open source framework of intelligent optimization methods.

@@

The experimental results presented in the paper, obtained on a series of GKLS-HC test problems, demonstrate the reliability of global search and the efficiency of the developed algorithm. By adjusting the parameter $\alpha$, it was possible to achieve the same speed of GSA-N operation as the basic GSA.

@@

Основным результатом являются эксперименты с модельными задачами настройки гиперпараметров, в которых возникают неопределенные значения оптимизируемой метрики качества. В этих задачах существуют недопустимые сочетания гиперпараметров, при которых настраиваемый метод возвращает бесконечное значение целевой функции.
Так была успешно проведена настройка классического алгоритма классификации LinearSVC. АГП-Н успешно решил поставленную задачу, найдя лучшее сочетание гиперпараметров по сравнению со стандартными алгоритмами из библиотеки scikit-learn.
Также был настроен метода предсказания значений временного ряда на базе фреймворка FEDOT. В ходе этого эксперимента также было проведено сравнение АГП-Н с алгоритмами настройки из другого широко известного фреймворка Optuna. АГП-Н показал сравнимый результат (в терминах полученного значения целевой метрики), значительно обогнав Optuna по времени, затраченному на решение задачи.

>>>>

The paper discusses  the problem of finding the best combination of hyperparameters of machine learning and artificial intelligence methods. In many cases, the efficiency (in a given metric) of the resulting solution for different values of hyperparameters can be quite different.
In such problems, a significant issue is the potential for incorrect operation of the investigated artificial intelligence and machine learning methods within certain (a priori unknown) subregions of the hyperparameters search domain. Furthermore, the computational complexity of tuning makes manual or exhaustive search inappropriate. These characteristics have required the development of various intelligent automatic hyperparameter optimization methods.
%В таких задачах актуальной является проблема некорректной работы исследуемых artificial intelligence (AI) and machine learning (ML) methods в некоторых (заранее неизвестных) подобластях области изменения гиперпараметров. При этом вычислительная сложность настройки делает невозможным использование ручного или полного перебора. Данные особенности определили необходимость разработки различных методов интеллектуального автоматического подбора гиперпараметров.
From a mathematical point of view, such a task can be represented as the problem of finding a global minimum of a function, given in the form of a ``black box'' and computable only in some part of the search domain. In this case, each computation of the objective function value at some point of the feasible domain may require significant computing resources. The objective function is assumed to satisfy the Lipschitz condition. The existence of subdomains where the objective function is undefined can be interpreted as the existence of some hidden, a priori unknown constraints of the problem. 
The authors propose an approach to solving this type of problem, which is an extension of the information-statistical global search algorithm (GSA) and takes into account the presence of undefined values of the objective function at some points. 
The algorithm partitions the search space with trial points and evaluates the characteristics of subregions based on the objective function values computed at their boundaries.  If the function value at a point is unknown, the algorithm employs an estimate for this value, considering the size of the subregion under investigation.
%В рамках алгоритма происходит разбиение области поиска точками испытаний и оцениваются характеристики подобластей на основе значений целевой функции, вычисленных на их границах. В случае отсутствия информации о значении функции в точке в алгоритме используется оценка этого значения, учитывающая размер исследуемой подобласти.
To minimize the number of redundant trials in subdomains where the function is not defined, the method parameter $\alpha$ was used that regulates the number of trial points in regions of non-computability.

@@

The article provides a detailed description and a flowchart of the operation of the proposed search algorithm.
The implementation of the global search algorithm for the case of a not everywhere computable objective function (GSA-N) was based on the iOpt open source framework of intelligent optimization methods.

@@

The experimental results presented in the paper, obtained on a series of GKLS-HC test problems, demonstrate the reliability of global search and the efficiency of the developed algorithm. By adjusting the parameter $\alpha$, it was possible to achieve the same performance of GSA-N operation as the basic GSA.

@@

Experiments were also conducted with hyperparameter tuning problems where undefined values of the quality metric arise. In these problems, certain hyperparameter combinations caused the method to return infinite values for the objective function. The LinearSVC classification algorithm was successfully tuned. GSA-N effectively solved the problem, identifying a better hyperparameter combination compared to standard scikit-learn algorithms. A time series prediction method from the FEDOT framework was also configured. During this experiment, GSA-N was compared to tuning algorithms from the popular Optuna framework. GSA-N achieved comparable performance in terms of the obtained the target metric value, significantly surpassing Optuna in solution time.
%Также проведены эксперименты с модельными задачами настройки гиперпараметров, в которых возникают неопределенные значения оптимизируемой метрики качества. В этих задачах существовали сочетания гиперпараметров, при которых настраиваемый метод возвращает бесконечное значение целевой функции. Была успешно проведена настройка классического алгоритма классификации LinearSVC. GSA-N успешно решил поставленную задачу, найдя лучшее сочетание гиперпараметров по сравнению со стандартными алгоритмами из библиотеки scikit-learn. Также был настроен метод предсказания значений временного ряда из фреймворка FEDOT. В ходе этого эксперимента было проведено сравнение GSA-N с алгоритмами настройки из другого широко известного фреймворка Optuna. АГП-Н показал сравнимый результат (в терминах полученного значения целевой метрики), значительно обогнав Optuna по времени, затраченному на решение задачи.
